Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2024-06-23 EvolutionaryRescue1.tex
revised after UBD comments

@@

% Title page
\title{Modeling the contribution of aneuplouidy to evolutionary rescue following anticancer drug exposure}
% Authors
\renewcommand\Affilfont{\small}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways normally targeted by the drugs. 
Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model how aneuploidy impacts the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability of a tumor to survive drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells. Additionally, we investigate the effect of the fraction of pre-existent aneuploid cells on the probability of evolutionary rescue  following therapy.  We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment, if such evolutionary rescue has occurred.
We propose that aneuploidy can play an important role in the relapse of smaller secondary tumors.
% TODO missing something here... ok
\end{abstract}

@@

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anticancer agents such as vemurafenib. Reversine-treated cells had a higher proliferation rate following drug exposure compared to wildtype cancer cells, due to selection of specific beneficial karyotypes.
 Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during chemotherapy treatment, despite having lower fitness before the onset of chemotherapy.
One proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}. Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene}, and elevated levels of DNA damage repair due to higher basal level of DNA damage \citep{zerbib2023human}.

An important aspect of aneuploidy is that the rate with which cells become aneuploid, that is the rate of chromosome mis-segregation, is several orders of magnitude higher than mutation rates \citep{bakker2023predicting}. Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster compared to acquiring a mutation. Moreover several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome mis-segregation rate is not compatible with cell survival and proliferation.

@@

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.  Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} who tested the predictions of their model with data from experiments in yeast and bacteria.

Here we study evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We assume that the affected cancer is melanoma, the chemotherapeutic agent is vemurfenib and that resistance is conferered by either trisomy in Chromosome 2 or trisomy in Chromosome 6. We estimate the effect of aneuploidy on the probability of evolutionary rescue when it acts as an evolutionary stepping stone (i.e. intermediary) towards resistant cancer cells. We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. Given that aneuploidy is present in many tumors already before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy. We find that aneuploidy increases the \emph{window of opportunity} when evolutionary rescue is able to prevent extinction.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. 
We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to wildtype cells in a drug-free environment.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

\paragraph{Stochastic simulations.} 
Simulations are performed using the \emph{Gillespie stochastic simulation algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.

@@

 Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}. Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as a \emph{stepping stone} for the generation of the mutant that rescues the cancer cell population.

We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division. 
\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division (for $u\ll10^{-3}$ no adaptation occurs and for $u\gg10^{-2}$ cells will acquire lethal nullisomies where both pairs of chromosomes are lost). \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we select $u=10^{-2}$ per cell division. For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. As a result, the fraction of cancer cells with aneuploidy of interest is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy of interest).

@@

All of the parameters discussed above are shown in \Cref{table1}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if wildtype cells die rapidly (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.  
In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs. \emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.

@@

For large tumors ($N \gg N_m^*$), the mean evolutionary rescue time (\cref{eq:AsymptoticTimeRules}) is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$). % I removed the mention of exponential distribution as it was not explained and I dont see why you need it here
 Increasing the per division mutation rate, $v$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
Finally, increasing the tumor size leads to shorter mean rescue time, as there are more wildtype cells that can mutate to become resistant.

Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to have the beneficial aneuploidy even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation. We calculated the mean evolutionary rescue time with standing genetic variation $\tilde{\tau}_a$ (\cref{meantimet2SGV}) and compared our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time. % WHAT is the purpose of this sentence? ok

In Appendix \ref{sec:appendix_distribution_time} we calculated the probability that a rescue mutation has been generated by time $t$. This allows us to examine whether aneuploidy accelerates or delays evolutionary rescue. We find that anenploidy  accelerates evolutionary rescue after $1/\Delta_a\approx100$, days after which no more rescue mutations are generated through direct mutation (\Cref{cdffig}A). This shows that rescue mutations are generated in aneuploid cells at latter timescales then in sensitive cells and, thus, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.   % WHAT is the purpose of this sentence? ok

@@

We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, and therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
% Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this? ok

@@

The detection time $\tau_a^{M}$ can be defined as the time for the tumor size to reach detection threshold $M$. We derive the mean detection time for $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}. We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{M}$ is negligible when compared to the scenario where the detection size is equal to the initial population size $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$). % WHAT DOES THAT MEAN? ok
However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}). Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors compared to the mean recurrence time, that is, the resistant tumor can be detected before reaching its initial size.

Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with drug therapy usually following after (i.e., adjuvant therapy). % WHAT DOES "varying" mean here? ok
Therefore, since only undetected secondary tumors are present at the time of treatment with anti-cancer drug, we lack data on the size of the tumors and cannot compare the empirical distributions to our predictions. However, we expect  the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data on cancer relapse.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}

We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size towards potential extinction.
In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths. In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow. In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

@@

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}). Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in large tumors in which the number of cells is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation in sensitive cells before these cells become extinct (\Cref{rescue_prob}). 
However, aneuploidy can have a crucial role in the evolutionary rescue of small tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors, which are too small to be detected and for which adjuvant chemotherapy is employed to prevent  cancer relapse (since we assume that the tumor is detected as soon as possible, large tumors are more likely to be primary tumors and small (undetected) tumors are secondary tumors). These secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.

@@

We hypothesized that \emph{standing genetic variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. We found that a drug or combination of drugs that reduce the wildtype growth rate and do not significantly increase the chromosome missegregation rate are more likely to cause evolutionary rescue through direct mutation (\cref{eq:Nsgv}). Furthermore, we found that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}). If the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de-novo} aneuploidy. In this scenario, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).

@@

\paragraph{Experimental future direction}
Our model predictions could be tested by experiments \citep{martin2013probability}. For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Then, these tumors may be exposed to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}. 
Cell density can be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present. We can then compare the results of the experiments to the predictions of our model to see if tumors with initial size bellow the threshold \cref{eq:N_a} are more likely to become extinct due to the drug.

%Additionally, our model predictions can be tested with data from patients (e.g. tumor karyotype). Given the type of cancer and how beneficial aneuploidy is,  our model can offer predictions about the probability for cancer relapse after treatment which can be tested against patient relapse data.

>>>>

% highlight
\newcommand{\hl}[1]{\colorbox{yellow!50}{$\displaystyle#1$}}

@@

% Title page
\title{
	Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs 
}
% Authors
\renewcommand\Affilfont{\small}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways normally targeted by the drugs. 
Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model how aneuploidy impacts the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells. Additionally, we investigate the effect of the fraction of pre-existent aneuploid cells on the probability of evolutionary rescue.  We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment, if such evolutionary rescue has occurred.
We propose that aneuploidy can play an important role in the relapse of smaller secondary tumors.
% TODO missing something here... ok
\end{abstract}

@@

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anti-cancer drugs such as vemurafenib. Reversine-treated cells had a higher proliferation rate following drug exposure compared to wildtype cancer cells, due to selection of specific beneficial karyotypes.
Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during drug treatment, despite having lower fitness before the onset of treatment.
One proposed mechanism through which aneuploidy is able to confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}. Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene}, and elevated levels of DNA damage repair due to higher basal level of DNA damage \citep{zerbib2023human}.

An important aspect of aneuploidy is that the rate with which cells become aneuploid, that is the rate of chromosome mis-segregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}. Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster compared to acquiring a mutation. Moreover several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome mis-segregation rate is not compatible with cell survival and proliferation.

@@

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.  Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability} who tested the predictions of their model with data from experiments in yeast and bacteria.

Here we study evolutionary rescue after a sudden environment change caused by the initiation of anti-cancer drug treatment. 
We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}.
% TODO explain distinction in a sentence
We estimate the effect of aneuploidy on the probability of evolutionary rescue of the tumor. When aneuploidy provides resistance to the drug, it can directly rescue the tumor. However, when anueploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background. 
As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to durg therapy, but it is unknown how often aneuploidy provides tolerance and thereby acts as a stepping stone.
We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. 
Given that aneuploidy is present in many tumors already before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy. We find that aneuploidy increases the \emph{window of opportunity} when evolutionary rescue is able to prevent extinction.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: sensitive, $w_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We consider a range of possible values for $\Delta_a$, finding three distinct cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.

We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.

See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

\paragraph{Stochastic simulations.} 
Simulations are performed using the \emph{Gillespie stochastic simulation algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: sensitive, aneuploid, and mutant. 
The sensitive population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.

@@

Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. 
For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the cancer cell population (note that this mutant will occur on the background of an aneuploid genotype).

{\color{red} The missegregation rate according to \citet{bakker2023predicting} is between $10^{-3}-10^{-2}$ per chromosome per cell division, with the most likely value being $6.2\times10^{-3}$ per chromosome per cell division.} 
\citet{ippolito2021gene} observed that trisomy in Chr 2 and Chr 6 are most likely to confer increased resistance against the anti-cancer drug vemurafenib for A375 cells. 
We assume each of these trisomies are formed at the most likely rate, and as a result, we use $u=10^{-2}$ per cell division. For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
We estimate $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. Thus, our estimate for the fitness cost $s$ is $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. Therefore, the estimated fraction of cancer cells with the beneficial aneuploidy is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$, that is, $0.28\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.

@@

All the parameters discussed above are shown in \Cref{table1}.

@@

Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced by the aneuploidy cost $s$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if wildtype cells die rapidly (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.  
In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs. \emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.

@@

For large tumors ($N \gg N_m^*$), the mean evolutionary rescue time (\cref{eq:AsymptoticTimeRules}) is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$). 
Increasing the per division mutation rate, $v$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
Finally, increasing the tumor size leads to shorter mean rescue time, as there are more wildtype cells that can mutate to become resistant.

Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to have the beneficial aneuploidy even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation. We calculated the mean evolutionary rescue time with standing genetic variation $\tilde{\tau}_a$ (\cref{meantimet2SGV}) and compared our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.

In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a rescue mutation has been generated by time $t$. This allows us to examine whether aneuploidy promotes or delays evolutionary rescue. 
We find that anenploidy promotes evolutionary rescue after $1/\Delta_a\approx100$ days, at a time where no more rescue mutations are generated through mutations in sensitive cells (\Cref{cdffig}A). Thus, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue. This can have a counter-intuitive outcome: conditioned on rescue of the tumor, tumors rescued by aneuploid cells may acquire rescue mutations later than those rescued by sensitive cells.

@@

We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, and therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.

@@

The detection time $\tau_a^{M}$ can be defined as the time for the tumor size to reach detection threshold $M$. We derive the mean detection time for $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}. We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{M}$ is negligible when compared to the scenario where the detection size is equal to the initial population size $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$). 
However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}). Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors compared to the mean recurrence time, that is, the resistant tumor can be detected before recovering back to its initial size.

\section*{Discussion}

We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size towards potential extinction.
In this scenario, the tumor can be ``evolutionary rescued'', or escape extinction, via two paths.
In the direct path, a sensitive cell acquires a mutation or aneuploidy that confers resistance that allows it to rapidly grow.
In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

@@

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
When the number of cells in the tumor is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$), a resistance mutation will occur in sensitive cells before these cells become extinct (\Cref{rescue_prob}). Therefore, large tumors are certain to be rescued with or without aneuploidy.
However, anti-cancer drugs are often used as adjuvant therapy after resection, in which case the number of cells in the tumor may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and therefore in cancer recurrence.
{\color{red} Indeed, secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.}

@@

We hypothesized that standing genetic variation (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. We found that a drug that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is likely to lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather then through \emph{de-novo} aneuploidy. However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.

@@

\paragraph{Experimental validation}
Our model predictions could be tested by experiments. For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes. Then, the extinction or survival of these tumors can be monitored during exposure to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}. 
We can then compare the results of these experiments to predictions of our model to see if tumors with initial size below the threshold \cref{eq:N_a} are more likely to become extinct due to drug exposure.
